Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-08 latex/main.tex
Updates from Overleaf

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
(1) RecombinHunt \cite{Alfonsi2024},
(2) CovRecomb \cite{Li2024}, and
(3) rebar (https://github.com/phac-nml/rebar).
These non-phylogenetic methods use the same underlying approach: 
they scan candidate recombinant sequences for mutations associated with different 
Pango lineages.

To evaluate the concordance of the methods in identifying the parental Pango lineages
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study;
for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods);
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods).
For each method, we determined whether the inferred parental lineages and breakpoint intervals are
concordant with those identified by the community (STAR Methods).
We find that the methods performed well overall,
with no significant difference in concordance among them
(Table~\ref{tab:x_method_concordance}; p-value = XX, chi-square test),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar.
Of all the ``robust'' recombination events detected by sc2ts,
167 (43\%) were also detected as recombinants by rebar (STAR Methods).
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we find that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
lower distance between their Pango identity (Figure
\ref{fig:averted_mutations}C).

\begin{table}
\caption{Concordance among methods in detecting Pango X lineages.}
\label{tab:method_concordance}
\begin{tabular}{llll}
\hline
\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
\hline
Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
\hline
RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
\hline
RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
\hline
CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
\hline
\end{tabular}
\end{table}

>>>>

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
(1) RecombinHunt \cite{Alfonsi2024},
(2) CovRecomb \cite{Li2024}, and
(3) rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach: 
they scan candidate recombinant sequences for mutations associated with different Pango lineages 
without using a phylogeny.

To evaluate the concordance of the methods in 
identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
we focus on the Pango Xs descending from the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took available detection results for these Pango Xs reported in the original study;
for CovRecomb, we summarized the results from the original study (STAR Methods);
For each Pango X, we determined whether the inferred parental lineages and breakpoint intervals 
are concordant with those suggested by the community (STAR Methods).
We found that the methods performed well overall, 
with no evidence for significant difference in the concordance rate among them (Table \ref{tab:x_method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

Furthermore, we compared the sensitivity of sc2ts and rebar.
We ran rebar on the ``robust'' recombination events detected by sc2ts (STAR Methods), 
and found that 167 (43\%) of them were also detected as recombinants by rebar.
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we found that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by 
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages. 
The Pango X lineages analyzed here are associated with the Type 1 recombination events (related to Table 1).
For RecombinHunt and CovRecomb, the concordance rates are calculated 
based on the Pango Xs for which results are available from the original studies.
Also, for RecombinHunt, there are two sets of detection results, 
one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.

}
\label{tab:method_concordance}
\begin{tabular}{llll}
\hline
\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bases)} \\ [1ex]
\hline
Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
\hline
RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
\hline
RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
\hline
CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
\hline
\end{tabular}
\end{table}
